---
title: "Biomass Mortality Analysis"
author: "breanna and Caroline S."
date: "2026-07-13"
output: 
  html_document:
    theme: cerulean
    code_folding: hide
    highlight: monochrome
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: yes
    smooth_scroll: yes
    number_sections: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# read in packages

library(janitor) #helps clean data, idk if it does anything else
library(tidyverse) #helps clean and manipulate data
library(here) #helps read in data 
library(dplyr) #helps with data analysis
library(stargazer) #helps make pretty tables to show data in knitted html or to save as png
library(lubridate) #helps R recognize dates
library(kableExtra)
```

## Introduction:
how
This code aims to synthesize and visualize data collected on Biomass (kg/m2) and Mortality of plants in the Elings Park Restoration Project. The Santa Barbara Botanic Garden, in partnership with Channel Islands Restoration, is transforming a 1-acre plot of land previously full of invasive weeds, into a better-functioning native habitat.

In this analysis, we explore the relationship between different weed treatments (sheet mulching and plastic tarping) and plant mortality and growth. 

```{r}
# read in the data
# Make sure you have the most updated data
elings_data <- read.csv(here("ElingsMonitoringData_2026July.csv")) %>% 
  clean_names() %>% 
  filter(species_id != "EMPTY") %>%  
  rename(width1_cm = width1_cm_widest_diameter, # changing column names to be less annoying
         width2_cm = width2_cm_perpendicular_to_w1_measurement) %>% 
  mutate(width1_m = width1_cm/100,  #making a width column in meters instead of cm
         width2_m = width2_cm/100,
         height2_m = height_cm) %>% 
  mutate(veg_type = case_when(species_id == "BOBA" ~ "herb",
                              species_id == "STPU" ~ "herb",
                              species_id == "BRCA" ~ "herb",
                              species_id == "CAMA" ~ "herb",
                              species_id == "COFI" ~ "herb",
                              species_id == "EPCA" ~ "herb",
                              species_id == "GRCA" ~ "herb",
                              species_id == "LECO" ~ "herb",
                              species_id == "SODO" ~ "herb",
                              species_id == "SPAM" ~ "herb",
                              species_id == "ACGL" ~ "shrub",
                              species_id == "AGSH" ~ "shrub",
                              species_id == "ARCA" ~ "shrub",
                              species_id == "ATLE" ~ "shrub",
                              species_id == "BAPI" ~ "shrub",
                              species_id == "BEEM" ~ "shrub",
                              species_id == "ENCA" ~ "shrub",
                              species_id == "ERFA" ~ "shrub",
                              species_id == "EUMI" ~ "shrub",
                              species_id == "HEWH" ~ "shrub",
                              species_id == "ISME" ~ "shrub",
                              species_id == "LEGI" ~ "shrub",
                              species_id == "LYCA" ~ "shrub",
                              species_id == "MAFA" ~ "shrub",
                              species_id == "PEAR" ~ "shrub",
                              species_id == "RHCR" ~ "shrub",
                              species_id == "RHIN" ~ "shrub",
                              species_id == "SAAP" ~ "shrub")) %>% 
  mutate(mort_status = case_when(mortality_status == "A" ~ "1",
                                      mortality_status == "D" ~ "0"))

class(elings_data$mortality_status)

elings_data$mort_status <- as.numeric(elings_data$mort_status)

class(elings_data$mort_status)

elings_data$date_lbd <- parse_date_time(elings_data$date, "ymd") 

class(elings_data$date_lbd)

# making sure Date is "Date" not "Character"
elings_data$date <- as.Date(elings_data$date)

class(elings_data$date)

elings_data <- elings_data %>% 
  filter(date > "2026-02-01")

# could create function to calculate biomass here vs Excel

```


# Grouping
```{r}

# group by plot and sum biomass 
biomass_plot <- elings_data %>% 
  group_by(plot_id,
           treatment) %>% 
  summarise(biomass_sum = sum(biomass_kg_m2))
# adding sd and se
biomass_plot <- elings_data %>% 
  group_by(plot_id, treatment) %>% 
  summarise(biomass_sum = sum(biomass_kg_m2),
            biomass_avg = mean(biomass_kg_m2, na.rm = TRUE),
            sd = sd(biomass_kg_m2, na.rm = TRUE),
            n = n(),
            se = sd /sqrt(n))




# group by plot and sum mortality
mort_plot <- elings_data %>% 
  group_by(plot_id,treatment) %>% 
  summarise(alive_plants = sum(mort_status, na.rm = TRUE))
#adding proportion
mort_plot <- elings_data %>%
  group_by(plot_id, treatment) %>%
  summarise(alive_plants = sum(mort_status, na.rm = TRUE),
            total_plants = n(),
            alive_prop = mean(mort_status),
            alive_pct = alive_prop * 100,
            .groups = "drop")




# group by treatment and sum biomass
biomass_treatment <- elings_data %>% 
  group_by(treatment) %>% 
  summarise(biomass_sum = sum(biomass_kg_m2))
#adding sd and se
biomass_treatment <- elings_data %>%
  group_by(treatment) %>%
  summarise(biomass_sum = sum(biomass_kg_m2),
            biomass_avg = mean(biomass_kg_m2, na.rm = TRUE),
            sd = sd(biomass_kg_m2, na.rm = TRUE),
            n = n(),
            se = sd / sqrt(n))




# group by treatment and sum mortality
mort_treatment <- elings_data %>% 
  group_by(treatment) %>% 
  summarise(alive_plants = sum(mort_status, na.rm = TRUE))
#add pct
mort_treatment <- elings_data %>% 
  group_by(treatment) %>% 
  summarise(alive_plants = sum(mort_status, na.rm = TRUE),
            total_plants = n(),
            alive_prop = mean(mort_status),
            alive_pct = alive_prop *100)




#group by species biomass and mortality
species_data <- elings_data %>% 
  group_by(species_id) %>% 
  summarise(alive_pl = sum(mortality_status == "A"),
            dead_pl = sum(mortality_status == "D"),
            total = alive_pl + dead_pl,
            alive_prop = alive_pl / total,
            biomass = sum(biomass_kg_m2),
            biomass_per_pl = biomass / alive_pl)

#group by species biomass and mortality by treatment
species_data_treatment <- elings_data %>% 
  group_by(species_id, treatment) %>% 
  summarise(alive_pl = sum(mortality_status == "A"),
            dead_pl = sum(mortality_status == "D"),
            total = alive_pl + dead_pl,
            alive_prop = alive_pl / total,
            biomass = sum(biomass_kg_m2),
            biomass_per_pl = biomass / alive_pl)

```


# Data exploration
Looking at trends in mortality and biomass based on plot and treatment.
```{r}

#graphing biomass by plot
ggplot(data = biomass_plot, aes(x = plot_id, y = biomass_sum, fill = treatment)) +
  geom_col() +
  theme_bw() +
  labs(x = "Plot ID",
       y = "Total Biomass (kg/m²)") 
# average biomass by plot with error bars
ggplot(data = biomass_plot, aes(x = plot_id, y = biomass_avg, fill = treatment))+
  geom_col() +
  geom_errorbar(aes(ymin = biomass_avg -se, ymax = biomass_avg + se), width = 0.2) +
  theme_bw() +
  labs(x = "Plot ID",
       y = "Average Biomass (kg/m²)")




# graphing biomass by treatment
ggplot(data = biomass_treatment, aes(x = treatment, y = biomass_sum, fill = treatment)) +
  geom_col() +
  theme_bw() +
  labs(x = "Treatment",
       y = "Total Biomass (kg/m²)")
#average biomass by treatment with error bars
ggplot(data = biomass_treatment, aes(x = treatment, y = biomass_avg, fill = treatment)) +
  geom_col() +
  geom_errorbar(aes(ymin = biomass_avg - se, ymax = biomass_avg + se), width = 0.2) +
  theme_bw() +
  labs(x = "Treatment", 
       y = "Mean Biomass (kg/m²)")




# graphing mortality by plot
ggplot(mort_plot, aes(x = plot_id, y = alive_plants, fill = treatment)) +
  geom_col() +
  theme_bw() +
  labs(x = "Plot ID",
       y = "Total Number of Live Plants")
ggplot(mort_plot, aes(x = plot_id, y = alive_pct, fill = treatment)) +
  geom_col() +
  scale_y_continuous(limits = c(0, 100)) +
  theme_bw() +
  labs(x = "Plot ID",
      y = "% of Plants Alive")




# graphing mortality by treatment
ggplot(data = mort_treatment, aes(x = treatment, y = alive_plants, fill = treatment)) + 
  geom_col() +
  theme_bw() +
  labs(x = "Treatment",
       y = "Total Number of Live Plants")
ggplot(data = mort_treatment, aes( x = treatment, y = alive_pct, fill = treatment)) +
  geom_col() +
  scale_y_continuous(limits = c(0,100)) +
  theme_bw() +
  labs(x = "Treatment",
       y = "% of Plants Alive")

```


# Continued analysis
The following code examines the differences in biomass and mortality for species based on region, province, treatment, and plot. Mkaing tables.
```{r}

#calculating % plants alive in each plot

#filtering for only tarp treatment to see the total number of plants (both A and D)
tarp_filt <- elings_data %>% 
  filter(mortality_status == "D" | mortality_status == "A") %>% 
  filter(treatment == "tarp") %>% 
  count(mortality_status) %>%
            bind_rows(summarise(., across(where(is.numeric), sum),
                                   across(where(is.character), ~'Total'))) %>% 
  rename(plant_count = n)


#filtering for only mulch treatment to see the total number of plants (both A and D)
mulch_filt <- elings_data %>% 
  filter(mortality_status == "D" | mortality_status == "A") %>% 
  filter(treatment == "mulch") %>% 
  count(mortality_status) %>%
            bind_rows(summarise(., across(where(is.numeric), sum),
                                   across(where(is.character), ~'Total'))) %>% 
  rename(plant_count = n)


#filtering for mortality data and grouping by plot
plot_filt <- elings_data %>% 
  group_by(plot_id) %>% 
  summarise(alive_pl = sum(mortality_status == "A"),
            dead_pl = sum(mortality_status == "D"),
            total = alive_pl + dead_pl,
            alive_prop = alive_pl / total,
            area = sum(area_m2),
            biomass = sum(biomass_kg_m2)) %>% 
  mutate(treatment = case_when(plot_id == "A" ~ "Tarp", # adding a treatment column
                               plot_id == "B" ~ "Mulch",
                               plot_id == "C" ~ "Tarp",
                               plot_id == "D" ~ "Mulch",
                               plot_id == "E" ~ "Mulch",
                               plot_id == "F" ~ "Tarp",
                               plot_id == "G" ~ "Mulch",
                               plot_id == "H" ~ "Tarp"))

#stargazer(plot_filt, summary = FALSE, type = "text")

plot_filt %>% 
kable(caption = "Plot Mortality Status Data",
      col.names = c("Plot ID","# Alive Plants","# Dead Plants", "Total","Proportion of plants alive", "Area","Biomass kg/m2", "Treatment"),
      format = "html",table.attr = "style = 'width:70%'",
      digits = 2) %>% 
kable_styling(bootstrap_options = "striped")

region_filt <- elings_data %>% 
  group_by(region) %>% 
  summarise(alive_pl = sum(mortality_status == "A"),
            dead_pl = sum(mortality_status == "D"),
            total = alive_pl + dead_pl,
            alive_prop = alive_pl / total,
            dead_prop = dead_pl / total,
            biomass = sum(biomass_kg_m2))



# Stipa province, but not needed anymore 
#province_stpu <- elings_data %>% 
 # filter(province_info == "SB" | province_info == "IRV" | province_info == "SD") %>% 
  #filter(species_id == "STPU") %>% 
  #group_by(province_info) %>% 
  #summarise(alive_pl = sum(mortality_status == "A"),
   #         dead_pl = sum(mortality_status == "D"),
    #       total = alive_pl + dead_pl,
    #        alive_prop = alive_pl / total,
     #       biomass = sum(biomass_kg_m2),
      #      biomass_per_pl = biomass / alive_pl)

#province_stpu %>%
 # kable(caption = "Stipa Pulchra Province Data",
  #      col.names = c("Province Info", "# Alive Plants", "# Dead Plants", "Total", "Proportion of plants alive", "Total Biomass kg/m2", "Biomass kg/m2 per plant"),
   #     format = "html", table.attr = "style = 'width:80%;'",
    #    digits = 2) %>% 
  #kable_styling(bootstrap_options = "striped")

#stargazer(province_stpu, summary = FALSE, type = "text")  


# Province data for EPCA
province_epca <- elings_data %>% 
  filter(province_info == "SB" | province_info == "IRV" | province_info == "SD") %>% 
  filter(species_id == "EPCA") %>% 
  group_by(province_info) %>% 
  summarise(alive_pl = sum(mortality_status == "A"),
            dead_pl = sum(mortality_status == "D"),
            total = alive_pl + dead_pl,
            alive_prop = alive_pl / total,
            biomass = sum(biomass_kg_m2),
            biomass_per_pl = biomass / alive_pl)

province_epca %>% 
  kable(caption = "Epilobium Canum Province Data",
        col.names = c("Province Info", "# Alive Plants", "# Dead Plants", "Total","Proportion of plants alive", "Biomass kg/m2", "Biomass kg/m2 per plant"),
        format = "html", table.attr = "style = 'width:80%;'",
        digits = 2) %>% 
  kable_styling(bootstrap_options = c("hover", "condensed"))

```



# looking at trends in species data 
```{r}
## Species biomass sum
ggplot(data = species_data, aes(x = species_id, y = biomass)) +
  geom_col() +
  theme_bw() +
  labs(x = "Species", 
       y = "Total Biomass (kg/m2)")
## Species biomass sum by treatment
ggplot(data = species_data_treatment, aes(x = species_id, y = biomass, fill = treatment))+
  geom_col(position = position_dodge(width = 0.9)) +
  theme_bw() +
  labs(x = "Species",
       y = "Total Biomass (kg/m2)")


## Species proportion living plants
ggplot(data = species_data, aes(x = species_id, y = alive_prop)) +
  geom_col() +
  theme_bw() +
  labs(x = "Species",
       y = "Proportion of living plants")
## Species proportion of living plants by treatment
ggplot(data = species_data_treatment, aes(x = species_id, y = alive_prop, fill = treatment))+
  geom_col(position = position_dodge(width = 0.9))+
  theme_bw() +
  labs(x = "Species",
       y = "Proportion of Living Plants")

```




# plotting species data in a table
```{r}
## species table with info
species_data %>% 
  kable(caption = "Mortality and Biomass data by Species",
        col.names = c("Species", "# Alive Plants", "# Dead Plants", "Total # Plants","Proportion of plants alive", "Total Biomass kg/m2", "Biomass kg/m2 per plant"),
        format = "html", 
        table.attr = "style = 'width:80%;'",
        digits = 2,
        na.rm = TRUE) %>% 
  kable_styling(bootstrap_options = c("hover", "condensed"))
```




# Adding Error Bars to species comparisons
```{r}

## species biomass comparison
species_data <- elings_data %>%
  group_by(species_id) %>%
  summarise(alive_pl = sum(mortality_status == "A"),
            dead_pl = sum(mortality_status == "D"),
            total = alive_pl + dead_pl,
            alive_prop = alive_pl / total,
            biomass = mean(biomass_kg_m2, na.rm = TRUE),
            sd = sd(biomass_kg_m2, na.rm = TRUE),
            n = sum(!is.na(biomass_kg_m2)),
            se = sd / sqrt(n))




ggplot(species_data, aes(x = species_id, y = biomass)) +
  geom_col() +
  geom_errorbar(aes(ymin = biomass - se, ymax = biomass + se),
    width = 0.2) +
  theme_bw() +
  labs(x = "Species", y = "Average Biomass (kg/m²) per plant")



## species mortality comparison
species_data <- elings_data %>%
  group_by(species_id) %>%
  summarise(
    alive_pl = sum(mortality_status == "A"),
    dead_pl = sum(mortality_status == "D"),
    total = alive_pl + dead_pl,
    alive_prop = alive_pl / total,
    se = sqrt(alive_prop * (1 - alive_prop) / total),
    biomass = sum(biomass_kg_m2),
    biomass_per_pl = biomass / alive_pl
  )
ggplot(data = species_data, aes(x = species_id, y = alive_prop)) +
  geom_col() +
  geom_errorbar(aes(ymin = alive_prop - se,
                    ymax = alive_prop + se),
                width = 0.2) +
  theme_bw() +
  labs( x = "Species",
    y = "Proportion of living plants")
```
  
  
  
```


## Future plans

Future analysis will include:

- looking at percent of plants alive in each plot and treatment

- color = treatment to make graphs better

- analyzing average biomass per plant per plot

- average biomass per species per treatment

- compare san diego scrub species across treatments (bergaro, lycium, agave, amorpha, euphorbia) how are species that wouldn't normally be found at Elings
```{r}

```

- maybe comparing species that are from further south, further inland, or tried and true SB natives

- compare mortality between provinces within treatments

- rates of growth in future surveys

- does decomposition correlate to plant growth in any way?


The analysis will (hopefully) answer these questions:

- Do tarped plots or mulched plots promote more growth?

- Which treatment has the lowest mortality rate?

- Which species performed the best (in terms of growth and survival) in each plot and each treatment? Is one treatment better for some plants, while the other treatment is better for other plants?

- Does origin impact plant mortality or growth?

- Which stpu location grew the most? Which stpu location had the best survival rates?
