---
title: "Trend Across Time"
author: "breanna"
date: "2026-07-15"
output: 
  html_document:
    theme: cerulean
    code_folding: hide
    highlight: monochrome
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: yes
    smooth_scroll: yes
    number_sections: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# read in packages

library(janitor) #helps clean data, idk if it does anything else
library(tidyverse) #helps clean and manipulate data
library(here) #helps read in data 
library(dplyr) #helps with data analysis
library(stargazer) #helps make pretty tables to show data in knitted html or to save as png
library(lubridate) #helps R recognize dates
library(kableExtra)
library(ggplot2)
```


## Introduction
This code aims to synthesize and visualize data collected on Biomass (kg/m2) and Mortality of plants in the Elings Park Restoration Project. The Santa Barbara Botanic Garden, in partnership with Channel Islands Restoration, is transforming a 1-acre plot of land previously full of invasive weeds, into a better-functioning native habitat.

In this analysis, we explore the relationship between different weed treatments (sheet mulching and plastic tarping) and plant mortality and growth ACROSS TIME.




# Reading in the data
```{r}

elings_data <- read.csv(here("ElingsMonitoringData_all.csv")) %>% 
  clean_names() %>% 
  filter(species_id != "EMPTY") %>% 
  rename(width1_cm = width1_cm_widest_diameter, # changing column names to be less annoying
         width2_cm = width2_cm_perpendicular_to_w1_measurement) %>% 
  mutate(width1_m = width1_cm/100,  #making a width column in meters instead of cm
         width2_m = width2_cm/100,
         height2_m = height_cm) %>% 
  mutate(mort_status = case_when(mortality_status == "A" ~ 1,
                                      mortality_status == "D" ~ 0))


```


# Grouping data
```{r}

biomass_time_treatment <- elings_data %>% 
  group_by(date, treatment) %>% 
  summarise(biomass_sum = sum(biomass_kg_m2),
            mean_biomass = mean(biomass_kg_m2, na.rm = TRUE),
            sd = sd(biomass_kg_m2, na.rm = TRUE),
            n = n(),
            se = sd / sqrt(n),
            .groups = "drop")


biomass_time_plot <- elings_data %>% 
  group_by(date, plot_id) %>% 
  summarise(biomass_sum = sum(biomass_kg_m2),
            mean_biomass = mean(biomass_kg_m2, na.rm = TRUE),
            sd = sd(biomass_kg_m2, na.rm = TRUE),
            n = n(),
            se = sd / sqrt(n),
            .groups = "drop")


mortality_time_treatment <- elings_data %>% 
  group_by(date, treatment) %>% 
  summarise(alive_plants = sum(mort_status, na.rm = TRUE),
            total_plants = n(),
            alive_prop = alive_plants/total_plants,
            alive_pct = alive_prop * 100,
            .groups = "drop")


mortality_time_plot <- elings_data %>% 
  group_by(date, plot_id) %>% 
  summarise(alive_plants = sum(mort_status, na.rm = TRUE),
            total_plants = n(),
            alive_prop = alive_plants/total_plants,
            alive_pct = alive_prop * 100,
            .groups = "drop")

```


#Plotting Biomass Treatment Data
```{r}
## mean biomass over time, by treatment
ggplot(biomass_time_treatment,
       aes( x = date,
            y = mean_biomass,
            color = treatment,
            group = treatment)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_biomass - se, 
                    ymax = mean_biomass + se),
                width = 0.2) +
  theme_bw() +
  labs( x = "Monitoring Date",
        y = "Mean Biomass (kg/m2)",
        color = "Treatment")


## total biomass over time, by treatment
ggplot(biomass_time_treatment, 
       aes(x = date, 
           y = biomass_sum,
           color = treatment, 
           group = treatment)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 3) +
  theme_bw() +
  labs( x = "Monitoring Date",
        y = "Total Biomass (kg/m2)",
        color = "Treatment")


```


#Plotting Biomass Plot ID Data
```{r}

## mean biomass over time, by plot
ggplot(biomass_time_plot,
       aes( x = date,
            y = mean_biomass,
            color = plot_id,
            group = plot_id)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_biomass - se, 
                    ymax = mean_biomass + se),
                width = 0.2) +
  theme_bw() +
  labs( x = "Monitoring Date",
        y = "Mean Biomass (kg/m2)",
        color = "Plot")
        

## total biomass over time, by plot 
ggplot(biomass_time_plot,
       aes(x = date,
           y = biomass_sum,
           color = plot_id,
           group = plot_id)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 3) +
  theme_bw()+
  labs(x = "Monitoring Date",
       y = "Total Biomass (kg/m2)")
```


#Plotting Mortality Treatment Data 
```{r}

## total alive plants by treatment, over time
ggplot(mortality_time_treatment,
       aes(x = date,
           y = alive_plants,
           color = treatment, 
           group = treatment)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 3) +
  theme_bw() +
  labs(x = "Monitoring Date",
       y = "Total Alive Plants") +
  scale_y_continuous(limits = c(0,750))


## proportion of alive plants by treatment, over time
ggplot(mortality_time_treatment,
       aes(x = date,
           y = alive_prop,
           color = treatment, 
           group = treatment)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 3) +
  theme_bw() +
  labs(x = "Monitoring Date",
       y = "Proportion of Alive Plants") +
  scale_y_continuous(limits = c(0,1.0))



```


#Plotting Mortality Plot Data
```{r}


## total alive plants by plot, over time
ggplot(mortality_time_plot,### PLOT A HAS WEIRD DATA LOOK INTO IT
       aes(x = date,
           y = alive_plants,
           color = plot_id, 
           group = plot_id)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 3) +
  theme_bw() +
  labs(x = "Monitoring Date",
       y = "Total Alive Plants") +
  scale_y_continuous(limits = c(0,270))



## proportion of alive plants by plot, over time
ggplot(mortality_time_plot,
       aes(x = date,
           y = alive_prop,
           color = plot_id, 
           group = plot_id)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 3) +
  theme_bw() +
  labs(x = "Monitoring Date",
       y = "Proportion of Alive Plants") +
  scale_y_continuous(limits = c(0,1))

```

