---
title: "Southern Species Analysis"
author: "breanna"
date: "2026-07-29"
output: 
  html_document:
    theme: cerulean
    code_folding: hide
    highlight: monochrome
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: yes
    smooth_scroll: yes
    number_sections: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# read in packages

library(janitor) #helps clean data, idk if it does anything else
library(tidyverse) #helps clean and manipulate data
library(here) #helps read in data 
library(dplyr) #helps with data analysis
library(stargazer) #helps make pretty tables to show data in knitted html or to save as png
library(lubridate) #helps R recognize dates
library(kableExtra)
library(ggplot2)
```


## Introduction
This code aims to synthesize and visualize data collected on Biomass (kg/m2) and Mortality of plants in the Elings Park Restoration Project. The Santa Barbara Botanic Garden, in partnership with Channel Islands Restoration, is transforming a 1-acre plot of land previously full of invasive weeds, into a better-functioning native habitat.

In this analysis, we explore the relationship between different weed treatments (sheet mulching and plastic tarping) and plant mortality and growth between LOCAL and SOUTHERN species.



# read in data and clean names
```{r}
elings_data <- read.csv(here("ElingsMonitoringData_all.csv")) %>% 
  clean_names() %>% 
  filter(species_id != "EMPTY") %>% 
  
  rename(width1_cm = width1_cm_widest_diameter,
         width2_cm = width2_cm_perpendicular_to_w1_measurement) %>% 
  
  mutate(width1_m = width1_cm/100,
         width2_m = width2_cm/100,
         height2_m = height_cm) %>% 
  
  mutate(mort_status = case_when(mortality_status == "A" ~ 1,
                                      mortality_status == "D" ~ 0))



elings_data$date <- as.Date(elings_data$date)

class(elings_data$date)

elings_data <- elings_data %>% 
  filter(date > "2026-02-01")

```


#grouping data
```{r}
#southern species only
south_species <- elings_data %>%
  filter(region == "south") %>% 
  group_by(species_id) %>% 
  summarise(biomass_sum = sum(biomass_kg_m2),
            mean_biomass = mean(biomass_kg_m2, na.rm = TRUE),
            alive_plants = sum(mort_status, na.rm = TRUE),
            total_plants = n(),
            alive_pct = alive_plants/total_plants *100,
            .groups = "drop")



#southern and selected local species
south_local_species <- elings_data %>% 
  arrange(region, species_id, treatment) %>% 
  filter((region == "central" & species_id %in% c("ARCA", "BAPI", "BOBA", "BRCA", "ERFA", "LECO")) |
           region == "south") %>% 
  group_by(region, species_id, treatment) %>% 
  summarise(biomass_sum = sum(biomass_kg_m2),
            mean_biomass = mean(biomass_kg_m2, na.rm = TRUE),
            alive_plants = sum(mort_status, na.rm = TRUE),
            total_plants = n(),
            alive_pct = alive_plants/total_plants *100,
            .groups = "drop")


```


#plotting data
```{r}
ggplot(south_local_species,
       aes(x = species_id,
           y = mean_biomass,
           fill = treatment,
           group = treatment)) +
  geom_col() +
  theme_bw() +
  labs(x = "Species",
       y = "Mean Biomass (kg/m2")
      # color = "Treatment")
  



ggplot(south_local_species,
       aes(x = species_id,
           y = mean_biomass,
           fill = region)) +
  geom_col() +
  theme_bw() +
  labs(x = "Species",
       y = "Mean Biomass (kg/m2")
  


ggplot(south_local_species,
       aes(x = species_id,
           y = mean_biomass,
           fill = treatment,
           group = treatment)) +
  geom_col() +
  facet_wrap() +
  theme_bw() +
  labs(x = "Species",
       y = "Mean Biomass (kg/m2",
       color = "Treatment")





```


#groupping over time
```{r}
region_compare <- elings_data %>%
  group_by(date, region) %>% 
  summarise(biomass_sum = sum(biomass_kg_m2),
            mean_biomass = mean(biomass_sum),
            alive_plants = sum(mort_status, na.rm = TRUE),
            total_plants = n(),
            alive_pct = alive_plants/total_plants *100,
            .groups = "drop")

biomass_time_treatment <- elings_data %>% 
  group_by(date, treatment) %>% 
  summarise(biomass_sum = sum(biomass_kg_m2),
            mean_biomass = mean(biomass_kg_m2, na.rm = TRUE),
            sd = sd(biomass_kg_m2, na.rm = TRUE),
            n = n(),
            se = sd / sqrt(n),
            .groups = "drop")


```
